---
title: "Peer assessement 1"
author: "Taha Bassit"
date: "October 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lattice)

```

##Loading and processing the Data set
###data base is read and the date column is converted into dates for analysis purposes.

```{r}
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date)
str(activity)

names(activity)
```

##What is mean total number of steps taken per day?
###Calculate the total number of steps taken per day

```{r}
dailyActivity <- sapply(split(activity$steps, activity$date),sum, na.rm=TRUE)
```

###Make a histogram of the total number of steps taken each day


```{r}
hist(dailyActivity, main= "Total daily steps", xlab = "number of steps", ylab = "frequency")

```

###Mean and Median number of daily steps.


```{r}

mean_dailyActivity <- mean(dailyActivity)
median_dailyActivity <-median(dailyActivity)

print(mean_dailyActivity)
print(median_dailyActivity)
```

##What is the average daily activity pattern?
###plotting the interval activity time series

```{r}
intervalActivity <-sapply(split(activity$steps, activity$interval),mean, na.rm=TRUE)
plot(intervalActivity, xlab="5-min intervals", type="l",ylab = "average daily steps", main="average number of interval steps")
```

###Identifiying the interval with the maximum average number of steps taken

```{r}
max_intervalID <- which.max(intervalActivity)

names(max_intervalID)
```

##Imputing the missing values
###Calculate and report the total number of missing values in the dataset

```{r}
sum(is.na(activity$steps))
```

###Filling strategy: replacing NA values with interval means.
####1-generating the NA free vector

```{r}
StepsAverage <- aggregate(steps ~ interval, data = activity, FUN = mean)
fillNA <- numeric()
for (i in 1:nrow(activity)) {
    obs <- activity[i, ]
    if (is.na(obs$steps)) {
        steps <- subset(StepsAverage, interval == obs$interval)$steps
    } else {
        steps <- obs$steps
    }
    fillNA <- c(fillNA, steps)
}
```

####2-Generating a new database by replacing the steps column with the new NA free vector

```{r}
new_activity <- activity
new_activity$steps <- fillNA

```

###- Generating daily activities and plotting the new data

```{r}
dailyActivity2 <-sapply(split(new_activity$steps, new_activity$date),sum, na.rm=TRUE)
hist(dailyActivity2, main= "Total daily steps", xlab = "number of steps", ylab = "frequency")

```

###Calculating new means and medians

```{r}

mean_dailyActivity2 <- mean(dailyActivity2)
median_dailyActivity2 <-median(dailyActivity2)

print(mean_dailyActivity2)
print(median_dailyActivity2)
```


##Pattern difference in activity between weekdays and weekends
###Generating weekday information

```{r}

new_activity$day <-weekdays(new_activity$date)  #generating an expanded database with variable day
fill_weekday <-character()
                                    #loop to generate the weekend or weekday information based on day test
for(i in 1:nrow(new_activity)){
      obs <-new_activity[i,]
      if (obs$day=="Saturday" | obs$day=="Sunday"){
            wday <-"Weekend"
      }else{
            wday <-"Weekday"
      }
      fill_weekday <-c(fill_weekday,wday)
}

```

###Creating a new data base that contains weekday information

```{r}

activity2 <-new_activity
activity2$weekday <-fill_weekday
names(activity2)
str(activity2)

```

###Plotting

```{r}

dailysteps <- aggregate(steps ~ interval + weekday, data = activity2, mean)


xyplot(steps ~ interval | weekday, dailysteps, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Number of steps")

```


